\section{Change report}
\subsection{Current situation}

In the orginal house.py file in the simulation folder, there is a function "scipy.integrate.odeint"  that calculates the ordinary differential equations from the domestic dwelling model. The SciPy website states however that this function should be replaced with scipy.integrate.solveivp. 

The code of the original ODE calculation in the house.py is as follows:

\begin{lstlisting}

for i in range(len(t)-1):

err = SP_T[i+1] - Tair[i]
Qinst = err * kp
Qinst = np.clip(Qinst, 0, 7000)

if (T_outdoor[i]>= 15):
Qinst=0
else:
Qinst=Qinst

inputs = (T_outdoor[i], Q_internal[i], Q_solar[i], SP_T[i], Qinst, CF,
Rair_outdoor, Rair_wall, Cair, Cwall)
# print(i)
ts = [t[i], t[i+1]]
y = odeint(model, y0, ts, args=inputs)

Tair[i+1] = y[-1][0]
Twall[i+1] = y[-1][1]
# integral[i+1]         = y[-1][2]

# Adjust initial condition for next loop

y0 = y[-1]
\end{lstlisting}

The way the odeint function is used in the original file is not correct. It is used in a for loop, in which only 2 points in time are used as input. Using these 2 points, the output of the ODE is generated. In the next iteration of the loop, outputs are used as inputs for the next time step. This causes unwanted oscillations in the output.

Control of the heat source is also included in the for loop, however, it would be more beneficial to have the control part of the model inside the house model itself.

\subsection{Improved situation}

This routine can be improved by using the solveivp function from the Scipy library and removing the for loop.

The function has similar arguments, however, in the solveivp function, the y0 and time data are switched (see below).


\begin{lstlisting}
	#odeint example
	y = odeint(model, y0, ts, args=inputs)
	#solveivp example
	y = solve_ivp(model_buffervessel, [0, t[-1]], y0, args=inputs)
\end{lstlisting}

The args=inputs in the odeint function are single values. However, for the solveivp function an array is used.

By incorporating the changes, the code snippet from the original file (see top of this document) can be replaced with the snippet below.

\begin{lstlisting}
inputs = (T_outdoor, Q_internal, Q_solar, SP_T, CF,
Rair_outdoor, Rair_wall, Cair, Cwall, UAradiator, Crad, Cbuffervessel, cpwater)
y = solve_ivp(model_buffervessel, [0, t[-1]], y0, args=inputs)
\end{lstlisting}

\subsection{Suggested code for house.py}
